Finite-state morphological transducers for three Kypchak languages

### 2014-03-22 kipchak-lrec-2013/kipchak-paper.tex
playing with space and stuff
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@51104 72bbbca6-d526-0410-a7d9-f06f51895060

\subsection{Kazakh}
Kazakh /q{\symbl ɑ}z{\symbl ɑ}q/ is spoken primarily in Kazakhstan, and it is the national language, where it is co-official with Russian.  Large communities of native speakers also exist in China, neighbouring Central-Eurasian republics, and Mongolia. Ethnologue estimates
the total number of speakers to be around 8 million \citep{ethnologue}.

@@

The transducers were designed based on the Helsinki Finite State Toolkit \citep{hfst/2011} which is a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis.  It implements both the \texttt{lexc} formalism for defining lexicons, and the \texttt{twol} and \texttt{xfst} formalisms for modeling morphophonological rules.  It also supports other finite state transducer formalisms such as \texttt{sfst}.  This toolkit has been chosen as it -- or the equivalent XFST \cite{beesley2003} -- has been widely used for other Turkic languages, such as Turkish \citep{coltekin2010}, Crimean~Tatar~\citep{altintas2001}, Turkmen \citep{tantug2006}, and Kyrgyz \citep{washington2012}, and is available under a free/open-source licence.

@@

\begin{figure}
\centering
%\hspace{-35pt}
    \rotatebox{-90}{
\includegraphics[width=0.38\textwidth,bb=0 50 554 770]{graphs/kumyk-yoldash-coverage}
}
%~\\
%~\\

@@

\begin{table}
\begin{center}
\begin{tabular}{lrrr}
		\toprule
\multirow{2}{*}{\textbf{Part of speech}} & \multicolumn{3}{c}{\textbf{Number of stems}} \\ \cline{2-4}
                        & Kazakh & Tatar & Kumyk \\
		\midrule
		Noun & 2640 & 2795 & 2568 \\
		Verb & 1470 & 1143 & 386 \\
		Adjective & 754 & 816 & 219 \\
		Proper noun & 5701 & 5361 & 1443 \\
		Adverb & 171 & 177 & 63 \\
		Numeral & 63 & 63 & 44 \\
		Conjunction & 46 & 45 & 13 \\
		Postposition & 50 & 43 & 12 \\
		Pronoun & 32 & 28 & 17 \\
		Determiner & 39 & 34 & 9 \\
		\midrule
		Total: & 11224 & 10737 & 4845 \\
		\bottomrule
\end{tabular}
 \caption{Number of stems in each of the categories. The large number of noun stems as compared
    to other parts of speech in the Kumyk transducer can be explained by the relative ease of 
    categorising nouns as opposed to adjectives, adverbs and verbs. Adding a noun essentially involves
    choosing between loan word and native word. Adding stems from the other main categories 
    requires more in depth categorisation.}
 \label{table:stems}
\end{center}

\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{llrr}
\toprule
\textbf{Language} & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) \\
\midrule
\multirow{4}{*}{Kazakh} & Wikipedia & 25.6M & 85.61 $\pm$ 1.37 \\
	& News & 3.8M & 92.12 $\pm$ 2.72  \\
	& Religion & 851K & 92.49 $\pm$ 1.66 \\\cline{2-4}
	& Average & -- & 90.07 $\pm$ 1.91 \\
\midrule
\multirow{4}{*}{Tatar} & Wikipedia & 159K & 86.35 $\pm$ 2.17 \\
	& News & 5.2M & 89.75 $\pm$ 0.07 \\
	& Religion & 382K & 91.25 $\pm$ 2.55 \\\cline{2-4}
	& Average & -- & 89.12 $\pm$ 1.60 \\
\midrule
% news = kum.yoldash.2013.txt
% religion = kum.genesis.ibt.txt, kum.nt.ibt.txt
\multirow{4}{*}{Kumyk} & Wikipedia & -- & -- \\
        & News & 286K &  91.10 $\pm$ 0.86 \\
	& Religion & 227K & 92.47 $\pm$ 1.03 \\\cline{2-4}
	& Average & -- & 91.78 $\pm$ 0.94 \\
\bottomrule
\end{tabular}
 \caption{Corpora used for naïve coverage tests}
 \label{table:corpora}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
	\begin{tabular}{lrr}
	\toprule
		\textbf{Language} & \textbf{Precision (\%)} & \textbf{Recall (\%)} \\
	\midrule
		Kazakh & 98.61 &  57.98 \\
		Tatar & 95.03 & 85.65 \\
		Kumyk & 96.57 & 69.11 \\
	\bottomrule
	\end{tabular}
	\caption{Precision and recall presented as percentages}
	\label{table:precrecall}
\end{center}
\end{table}

>>>>

\subsection{Kazakh}
Kazakh /q{\symbl ɑ}z{\symbl ɑ}q/ is spoken primarily in Kazakhstan, where it is the national language and is co-official with Russian.  Large communities of native speakers also exist in China, neighbouring Central-Eurasian republics, and Mongolia. Ethnologue estimates
the total number of speakers to be around 8 million \citep{ethnologue}.

@@

The transducers were designed based on the Helsinki Finite State Toolkit \citep{hfst/2011} which is a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis.  It implements both the \texttt{lexc} formalism for defining lexicons, and the \texttt{twol} and \texttt{xfst} formalisms for modeling morphophonological rules.  It also supports other finite state transducer formalisms such as \texttt{sfst}.  This toolkit has been chosen as it -- or the equivalent XFST \citep{beesley2003} -- has been widely used for other Turkic languages, such as Turkish \citep{coltekin2010}, Crimean~Tatar~\citep{altintas2001}, Turkmen \citep{tantug2006}, and Kyrgyz \citep{washington2012}, and is available under a free/open-source licence.

@@

\begin{figure}
%\centering
%\hspace{-35pt}
%\hspace{-1em}
%    \rotatebox{-90}{
%\hspace{-15pt}
\hspace{-10pt}
\includegraphics[width=0.37\textwidth,bb=50 50 554 770,angle=270]{graphs/kumyk-yoldash-coverage}
%}
%~\\
%~\\

@@

\begin{table}
\begin{center}
\begin{tabular}{lrrr}
		\toprule
\multirow{2}{*}{\textbf{Part of speech}} & \multicolumn{3}{c}{\textbf{Number of stems}} \\ \cline{2-4}
                        & Kazakh & Tatar & Kumyk \\
		\midrule
		Noun & 2640 & 2795 & 2568 \\
		Verb & 1470 & 1143 & 386 \\
		Adjective & 754 & 816 & 219 \\
		Proper noun & 5701 & 5361 & 1443 \\
		Adverb & 171 & 177 & 63 \\
		Numeral & 63 & 63 & 44 \\
		Conjunction & 46 & 45 & 13 \\
		Postposition & 50 & 43 & 12 \\
		Pronoun & 32 & 28 & 17 \\
		Determiner & 39 & 34 & 9 \\
		\midrule
		Total: & 11224 & 10737 & 4845 \\
		\bottomrule
\end{tabular}
 \caption{Number of stems in each of the categories. The large number of noun stems as compared
    to other parts of speech in the Kumyk transducer can be explained by the relative ease of 
    categorising nouns as opposed to adjectives, adverbs and verbs. Adding a noun essentially involves
    choosing between loan word and native word. Adding stems from the other main categories 
    requires more in depth categorisation.}
 \label{table:stems}
\end{center}
\vspace{-2em}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{llrr}
\toprule
\textbf{Language} & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) \\
\midrule
\multirow{4}{*}{Kazakh} & Wikipedia & 25.6M & 85.61 $\pm$ 1.37 \\
	& News & 3.8M & 92.12 $\pm$ 2.72  \\
	& Religion & 851K & 92.49 $\pm$ 1.66 \\\cline{2-4}
	& Average & -- & 90.07 $\pm$ 1.91 \\
\midrule
\multirow{4}{*}{Tatar} & Wikipedia & 159K & 86.35 $\pm$ 2.17 \\
	& News & 5.2M & 89.75 $\pm$ 0.07 \\
	& Religion & 382K & 91.25 $\pm$ 2.55 \\\cline{2-4}
	& Average & -- & 89.12 $\pm$ 1.60 \\
\midrule
% news = kum.yoldash.2013.txt
% religion = kum.genesis.ibt.txt, kum.nt.ibt.txt
\multirow{4}{*}{Kumyk} & Wikipedia & -- & -- \\
        & News & 286K &  91.10 $\pm$ 0.86 \\
	& Religion & 227K & 92.47 $\pm$ 1.03 \\\cline{2-4}
	& Average & -- & 91.78 $\pm$ 0.94 \\
\bottomrule
\end{tabular}
 \caption{Corpora used for naïve coverage tests}
 \label{table:corpora}
\end{center}
\vspace{-1.5em}
\end{table}

@@

\begin{table}
\begin{center}
	\begin{tabular}{lrr}
	\toprule
		\textbf{Language} & \textbf{Precision (\%)} & \textbf{Recall (\%)} \\
	\midrule
		Kazakh & 98.61 &  57.98 \\
		Tatar & 95.03 & 85.65 \\
		Kumyk & 96.57 & 69.11 \\
	\bottomrule
	\end{tabular}
	\caption{Precision and recall presented as percentages}
	\label{table:precrecall}
\end{center}
\vspace{-1.5em}
\end{table}
